---
title: "Untitled"
output:
  pdf_document: default
  html_document: default
date: "2022-10-27"
---
```{r,echo=F}
#Initial environment setup
setwd("~/Desktop/GLM/Coursework/GLM-CW")
expend.df <- read.table("expenditure.txt")
library(tidyverse)
library(kableExtra)
library(gridExtra)
library(broom)
library(MASS)
```

```{r,echo=F}
#Data cleaning

#coding variables as factors
expend.df$house.ten <- factor(expend.df$house.ten,levels=c(1:3),labels=c("Public Rented","Private rented","Owned"))
expend.df$sex.hh <- factor(expend.df$sex.hh,levels=c(1,2),labels=c("Male","Female"))
expend.df$lab.force <- factor(expend.df$lab.force,levels=c(1:4),labels=c("Full time", "Part time","Unemployed","Inactive"))
expend.df$hh.size <- factor(expend.df$hh.size,levels=c(1:5),labels = c("1 person","2 people","3 people","4 people","5+ persons"))
expend.df$hh.adults <- factor(expend.df$hh.adults,levels = c(1:4), labels = c("1 adult","2 adults","3 adults","4+ adults"))

```
## Task 1:

#### Part 1:
```{r,echo=F}
#Histogram to examine the expenditure
hist(expend.df$expenditure,breaks = 20,
     xlab = "Total weekly household expenditure (GBP)",
     main = "Histogram of expenditure")

#Creating the 
#scatter plot to look at relationship between income and expenditure
p1<-ggplot(data=expend.df,aes(x=income,y=expenditure))+
  geom_point(size=0.2)+
  xlab("Gross weekly average household income (GBP)")+
  ylab("Total expenditure")+
  theme(axis.text=element_text(size=6),
        axis.title=element_text(size=6,face="bold"))

#boxplots
p2<-ggplot(data=expend.df,aes(y=expenditure,x=house.ten,fill=house.ten))+
  geom_boxplot()+
  xlab("Household tenure")+
  guides(fill=F)

#scale_fill_discrete(name="Household\ntenure")

p3<-ggplot(data=expend.df,aes(y=expenditure,x=sex.hh,fill=sex.hh))+
  geom_boxplot()+
  xlab("Sex of the household head")+
  guides(fill=F)
  
p4<-ggplot(data=expend.df,aes(y=expenditure,x=lab.force,fill=lab.force))+
  geom_boxplot()+
  xlab("Employment status")+
  guides(fill=F)

p5<-ggplot(data=expend.df,aes(y=expenditure,x=hh.size,fill=hh.size))+
  geom_boxplot()+
  xlab("Household size")+
  guides(fill=F)

p6<-ggplot(data=expend.df,aes(y=expenditure,x=hh.adults,fill=hh.adults))+
  geom_boxplot()+
  xlab("Number of adults in the household")+
  guides(fill=F)

grid.arrange(p1,p2,p3,p4,p5,p6,ncol=2,nrow=3)
```



Looking at the above histogram of expenditure above it shows that expenditure has a positively skewed distribution.
Investigating the relationship between household income and expenditure in the scatter plot above, it displays a clear positive association.
Then to investigate the other variables which are categorical I used boxplots to display differences in the expenditure across the different levels(categories). The differences shown in the plots are listed below:

 
-Different household tenures categories have different weekly expenditure amounts with public rented households having a noticeably lower expenditure compared to the other 2 household levels.
-Households with a female head have a lower weekly expenditure compared to male household heads
-Employment status appears to be associated with expenditure with unemployed households having a dramatically lower expenditure compared to households with full or part time employment.
-Household size generally appears to be positively correlated with expenditure, with larger households having a higher weekly expenditure (however 5+ households break this trend with the expenditure decreasing, this could be due to other variables).
Number of adults in the household also appears positively correlated with expenditure with households containing more adults having a higher expenditure.

I note that these differences may not be due to the variable mentioned but due to a third variable (a confounder) that affects both the categorical variable and expenditure (the response) which could be other variables in the dataset or ones that were not even included.


#### Part 2:
In all linear models mentioned for part 2 to 6, the response $y$ is the weekly household expenditure and the explanatory variable $x$ is weekly income and $i \in \{1,...,1200\}$ is the ith household in the dataset.

```{r,echo=F}
#Question 2
lm.income<-lm(expenditure~income,data = expend.df)
#summary(lm.income)
as.data.frame(tidy(lm.income)[,1:3])%>%
  kable(escape = F)%>%
  kable_styling(font_size = 15)
par(mfrow=c(2,2))
plot(lm.income)
```
The estimated coefficients for the fitted linear model $y_{i}=\beta_{0}+\beta_{1}x_{i}+\epsilon$  are displayed in the above table. It shows the estimated values of the intercept ($\beta_0$) and the estimated 'effect' of a 1 pound increase in income on expenditure $\beta_1$ and the associated standard errors.

Viewing the diagnostic plots for this model. The fitted vs residual plot shows the residuals are not equally spread around 0 with a greater spread of residuals above zero, meaning the assumption of linearity may not hold. Furthermore, the QQ normal plot has a u shape with the tails diverging heavily away from the straight line. This implies the normality assumption is invalid.

#### Part 3:
```{r,echo=F}
#Question 3
lm.income.sqd <- lm(expenditure~income+I(income^2),data = expend.df)
tidy(lm.income.sqd)[,1:3]%>%
  kable(escape = F)%>%
  kable_styling(font_size = 15)
par(mfrow=c(2,2))
plot(lm.income.sqd)

# lm.log.income.sqd <- lm(log(expenditure)~income+I(income^2),data = expend.df)
# summary(lm.log.income.sqd)
# plot(lm.log.income.sqd)
# 
# gamma.income.sqd <-glm(expenditure~income+I(income^2),data=expend.df,family="Gamma")
# summary(gamma.income.sqd)
# plot(gamma.income.sqd)
```
Fitting a new linear model $y_{i}=\beta_{0}+\beta_{1}x_{i}+\beta_2x^2_{i}+\epsilon$ with an added squared term of income gives the above estimated coefficients. Looking at the diagnostic plots, in the fitted vs residual plot the residual values increase as the fitted values increase thus the homoscedasticity assumption may not be valid. The tails in the normal QQ plot also still deviate heavily away from the straight line meaning the normality assumption is invalid.

#### Part 4:
```{r,echo=F}
#Question 4
lm.log.income<-lm(log(expenditure)~income,data = expend.df)
tidy(lm.log.income)[,1:3]%>%
  kable(escape = F)%>%
  kable_styling(font_size = 15)
par(mfrow=c(2,2))
plot(lm.log.income)
# #QQ plot much better but not sure about the residual vs fitted if the variance decreases as fitted value increases
```
Fitting the same model as in task 2 but with the response of expenditure logged $log(y_{i})=\beta_{0}+\beta_{1}x_{i}+\epsilon$ gives the above coefficients. The data appears to be heteroscedastic as the fitted vs residual plot points spread decreases as fitted values increases, furthermore the linear assumption also appears violated as the residuals appear to exhibit a slight negative quadratic shape. The QQ plot shows the residuals stay close to the straight line thus normality assumption is valid. 

#### Part 5:
```{r,echo=F}
lm.log.income.sqd <- lm(log(expenditure)~income+I(income^2),data = expend.df)
tidy(lm.log.income.sqd)[,1:3]%>%
  kable(escape = F)%>%
  kable_styling(font_size = 15)
par(mfrow=c(2,2))
plot(lm.log.income.sqd)
```
Fitting the model $log(y_{i})=\beta_{0}+\beta_{1}x_{i}+\beta_{2}x^2_{i}+\epsilon$ gives the estimated coefficients above. Viewing the diagnostic plots the fitted vs residuals plot shows a fairly even scattering of points around 0 and a linear trend, however the residuals are bunched to the very right of the plot and the variance of the residuals appears to decrease as the fitted values increase, this could be to important variables being left out of the model (MAY NEED TO CHANGE THIS). The non-constant variance assumption being violated is further backed up by the none straight line in the scale location plot. The QQ plot shows the points stay close to the straight line meaning the normality assumption is valid.

#### Task 6
```{r}
#Comparing AIC
AIC(lm.income)
AIC(lm.income.sqd)
AIC(lm.log.income)
AIC(lm.log.income.sqd)
#Comparing R squared
glance(lm.income)$r.squared
glance(lm.income.sqd)$r.squared
glance(lm.log.income)$r.squared
glance(lm.log.income.sqd)$r.squared
#Comparing adjusted R squared
glance(lm.income)$adj.r.squared
glance(lm.income.sqd)$adj.r.squared
glance(lm.log.income)$adj.r.squared
glance(lm.log.income.sqd)$adj.r.squared
```

The 4th model $log(y_{i})=\beta_{0}+\beta_{1}x_{i}+\beta_{2}x^2_{i}+\epsilon$ has the lowest AIC value of *1528.062* and the highest adjusted R squared value of *0.4635055*. Furthermore, looking at the diagnostic plots the normality assumption is valid and the fitted vs residual plot although not 'perfect' displays a more even scattering of points than that of the same plot for the other models. Therefore for all these reasons the 4th model is my preferred model to describe the relationship between expenditure and income.


```{r}
tidy(lm.log.income.sqd)

mean(expend.df$income)
```
Income has a positive association with expenditure with a 1 pound increase in household income increasing household expenditure by $exp(0.0032-0.0000029x-0.0000015)-1$x100%. Therefore for a household earning the mean income of £512 a £1 increase in income increases expenditure by approximately 0.17%. Using a more likely realistic increase in weekly income of £20 for a household earning £512 gives an increase of $exp(0.016-0.000015*20-0.000036)-1$x100% = 1.6%.
#### Task 7
For interactions I decided to only consider 2 way interactions that could be plausible for example income may have a different effect on expenditure depending on sex of household head. Furthermore the number of adults or household tenure type may impact the effect of income on expenditure as.... above I investigated if there were differences when sub categorising the categories to see if there where plausible interactions between factor variables, however none showed any significant differences between the different cohorts (initial factors)

#### Model selection
For this model selection process I decided to only consider 2 way interactions as higher order interactions make interpretation of the model much more complicated.
Furthermore, when considering which interactions to include I first created plots to examine if there are any clear interactions between income and the other covariates. Considering plausible interactions between 2 factor covarites an interaction between household size and number of adults made sense?
and labour force with either household size or number of adults in the household.

household size and number of adults (number of adults may change effect of household size)
lab.force and hh.size or hh.adult as families who are unemployed may have different spending habits be more frugal then a singluar household that is unemployed 
 INVESTIGATE HOW SIMILAR EACH VARIABLE IS!
none made much logical sense e.g. a household female head would have a different effect on expenditure to a male head depending on the number of people in the household. Thus didn't include any factor:factor interactions in the model.

```{r}
#interaction between sex and income
interaction.continous <-function(var){
  ggplot(data=expend.df,aes(x=income,y=expenditure,colour=.data[[var]]))+
    geom_point(size=0.1)
}


int1<-interaction.continous("house.ten")
int2<-interaction.continous("sex.hh")
int3<-interaction.continous("lab.force")
int4<-interaction.continous("hh.size")
int5<-interaction.continous("hh.adults")

grid.arrange(int1,int2,int3,int4,int5,ncol=2,nrow=3)
int1
int2
int3
int4
int5
```
Looking at the scatter plots each factor for the plots does appear to have a different gradient thus an interaction could be present. Therefore for the model selection process I will start with a full model with all covariates and the pairwise interactions between income and each of the factor variables. Then using AIC I remove the coefficient that gives the biggest increase in AIC until no increase in AIC can be attained. Using this step method I will come to my first potential final model. 

```{r}
lm.log.full<-lm(log(expenditure)~income+I(income^2)+income*sex.hh+income*hh.adults+income*house.ten+lab.force*income+hh.size*income,data=expend.df)
step.method<- step(lm.log.full,trace=F)
tidy(step.method)
AIC(step.method)
glance(step.method)$adj.r.squared

#got rid of 2 interactions: income:sex and income:hhsize
AIC(lm.log.income.sqd)
AIC(step.method)
glance(lm.log.income.sqd)$adj.r.squared
glance(step.method)$adj.r.squared

summary(step.method)

anova(step.method)
```

Using AIC as the model selection criteria I removed 2 interactions income:sex and income:household size. Comparing this new model to the original the AIC is significantly lower and the adjusted R squarded is significantly higher. Also note that each variable is significant except for 


```{r}
plot(step.method)
```




```{r}
lm.log.full.no.int<-lm(log(expenditure)~income+I(income^2)+sex.hh+hh.adults+house.ten+lab.force+hh.size,data=expend.df)
summary(lm.log.full.no.int)
step.method<- step(lm.log.full.no.int,trace=F)

plot(lm.log.full.no.int)
```

Using AIC as the selection critera the best model without interactions is the full model.

Now I try adding interactions that couold make logical sense.

As a sex of the household head could impact the effect of income on expenditure (perhaps males are more likely to spend then save then females as they get more money or the reverse)

The number of adults my affect the affect of income as more adults from a higher income household may be more likely to spend more 

The type of household tenure may affect the

```{r}
#adding these into the model 1 by 1
new.lm.add.sex<-update(lm.log.full.no.int, .~.+income*sex.hh)
anova(lm.log.full.no.int,new.lm.add.sex)
summary(new.lm.add.sex)

AIC(lm.log.full.no.int)
AIC(new.lm.add.sex)

glance(lm.log.full.no.int)$adj.r.squared
glance(new.lm.add.sex)$adj.r.squared
#Not really much difference and note that interaction is not significant

new.lm.add.adults<-update(lm.log.full.no.int, .~.+income*hh.adults)
anova(lm.log.full.no.int,new.lm.add.adults)
summary(new.lm.add.adults)

AIC(lm.log.full.no.int)
AIC(new.lm.add.adults)

glance(lm.log.full.no.int)$adj.r.squared
glance(new.lm.add.adults)$adj.r.squared


new.lm.add.ten<-update(lm.log.full.no.int, .~.+income*house.ten)
anova(lm.log.full.no.int,new.lm.add.ten)
summary(new.lm.add.ten)

AIC(lm.log.full.no.int)
AIC(new.lm.add.ten)

glance(lm.log.full.no.int)$adj.r.squared
glance(new.lm.add.ten)$adj.r.squared


AIC(lm.log.full)
AIC(lm.log.full.no.int)
glance(lm.log.full.no.int)$adj.r.squared
glance(lm.log.full)$adj.r.squared
```

```{r}
# Attempting Box-Cox transformation to check whether distribution is better
estim_BoxCox <- boxcox(expenditure ~ income+sex.hh+hh.adults+house.ten+lab.force+hh.size+income:hh.adults+income:house.ten+income:lab.force,data =expend.df)
# Find optimal lambda
lambda <- estim_BoxCox$x[ which.max( estim_BoxCox$y ) ]
```


