---
title: "Untitled"
output:
  pdf_document: default
  html_document: default
date: "2022-10-27"
---
```{r,echo=F}
#Initial environment setup
setwd("~/Desktop/GLM/Coursework/GLM-CW")
expend.df <- read.table("expenditure.txt")
library(tidyverse)
library(kableExtra)
library(gridExtra)
library(broom)
library(MASS)
```

```{r,echo=F}
#Data cleaning

#coding variables as factors
expend.df$house.ten <- factor(expend.df$house.ten,levels=c(1:3),labels=c("Public Rented","Private rented","Owned"))
expend.df$sex.hh <- factor(expend.df$sex.hh,levels=c(1,2),labels=c("Male","Female"))
expend.df$lab.force <- factor(expend.df$lab.force,levels=c(1:4),labels=c("Full time", "Part time","Unemployed","Inactive"))
expend.df$hh.size <- factor(expend.df$hh.size,levels=c(1:5),labels = c("1 person","2 people","3 people","4 people","5+ persons"))
expend.df$hh.adults <- factor(expend.df$hh.adults,levels = c(1:4), labels = c("1 adult","2 adults","3 adults","4+ adults"))

```
## Task 1:

#### Part 1:
```{r,echo=F}
#Histogram to examine the expenditure
hist(expend.df$expenditure,breaks = 20,
     xlab = "Total weekly household expenditure (GBP)",
     main = "Histogram of expenditure")

#Creating the 
#scatter plot to look at relationship between income and expenditure
p1<-ggplot(data=expend.df,aes(x=income,y=expenditure))+
  geom_point(size=0.2)+
  xlab("Gross weekly average household income (GBP)")+
  ylab("Total expenditure")+
  theme(axis.text=element_text(size=6),
        axis.title=element_text(size=6,face="bold"))

#boxplots
p2<-ggplot(data=expend.df,aes(y=expenditure,x=house.ten,fill=house.ten))+
  geom_boxplot()+
  xlab("Household tenure")+
  guides(fill=F)

#scale_fill_discrete(name="Household\ntenure")

p3<-ggplot(data=expend.df,aes(y=expenditure,x=sex.hh,fill=sex.hh))+
  geom_boxplot()+
  xlab("Sex of the household head")+
  guides(fill=F)
  
p4<-ggplot(data=expend.df,aes(y=expenditure,x=lab.force,fill=lab.force))+
  geom_boxplot()+
  xlab("Employment status")+
  guides(fill=F)

p5<-ggplot(data=expend.df,aes(y=expenditure,x=hh.size,fill=hh.size))+
  geom_boxplot()+
  xlab("Household size")+
  guides(fill=F)

p6<-ggplot(data=expend.df,aes(y=expenditure,x=hh.adults,fill=hh.adults))+
  geom_boxplot()+
  xlab("Number of adults in the household")+
  guides(fill=F)

grid.arrange(p1,p2,p3,p4,p5,p6,ncol=2,nrow=3)
```



Looking at the above histogram of expenditure above it shows that expenditure has a positively skewed distribution.
Investigating the relationship between household income and expenditure in the scatter plot above, it displays a clear positive association.
Then to investigate the other variables which are categorical I used boxplots to display differences in the expenditure across the different levels(categories). The differences shown in the plots are listed below:

 
-Different household tenures categories have different weekly expenditure amounts with public rented households having a noticeably lower expenditure compared to the other 2 household levels.
-Households with a female head have a lower weekly expenditure compared to male household heads
-Employment status appears to be associated with expenditure with unemployed households having a dramatically lower expenditure compared to households with full or part time employment.
-Household size generally appears to be positively correlated with expenditure, with larger households having a higher weekly expenditure (however 5+ households break this trend with the expenditure decreasing, this could be due to other variables).
Number of adults in the household also appears positively correlated with expenditure with households containing more adults having a higher expenditure.

I note that these differences may not be due to the variable mentioned but due to a third variable (a confounder) that affects both the categorical variable and expenditure (the response) which could be other variables in the dataset or ones that were not even included.


#### Part 2:
In all linear models mentioned for part 2 to 6, the response $y$ is the weekly household expenditure and the explanatory variable $x$ is weekly income and $i \in \{1,...,1200\}$ is the ith household in the dataset.

```{r,echo=F}
#Question 2
lm.income<-lm(expenditure~income,data = expend.df)
#summary(lm.income)
as.data.frame(tidy(lm.income)[,1:3])%>%
  kable(escape = F)%>%
  kable_styling(font_size = 15)
par(mfrow=c(2,2))
plot(lm.income)
```
The estimated coefficients for the fitted linear model $y_{i}=\beta_{0}+\beta_{1}x_{i}+\epsilon$  are displayed in the above table. It shows the estimated values of the intercept ($\beta_0$) and the estimated 'effect' of a 1 pound increase in income on expenditure $\beta_1$ and the associated standard errors.

Viewing the diagnostic plots for this model. The fitted vs residual plot shows the residuals are not equally spread around 0 with a greater spread of residuals above zero, meaning the assumption of linearity may not hold. Furthermore, the QQ normal plot has a u shape with the tails diverging heavily away from the straight line. This implies the normality assumption is invalid.

#### Part 3:
```{r,echo=F}
#Question 3
lm.income.sqd <- lm(expenditure~income+I(income^2),data = expend.df)
tidy(lm.income.sqd)[,1:3]%>%
  kable(escape = F)%>%
  kable_styling(font_size = 15)
par(mfrow=c(2,2))
plot(lm.income.sqd)

# lm.log.income.sqd <- lm(log(expenditure)~income+I(income^2),data = expend.df)
# summary(lm.log.income.sqd)
# plot(lm.log.income.sqd)
# 
# gamma.income.sqd <-glm(expenditure~income+I(income^2),data=expend.df,family="Gamma")
# summary(gamma.income.sqd)
# plot(gamma.income.sqd)
```
Fitting a new linear model $y_{i}=\beta_{0}+\beta_{1}x_{i}+\beta_2x^2_{i}+\epsilon$ with an added squared term of income gives the above estimated coefficients. Looking at the diagnostic plots, in the fitted vs residual plot the residual values increase as the fitted values increase thus the homoscedasticity assumption may not be valid. The tails in the normal QQ plot also still deviate heavily away from the straight line meaning the normality assumption is invalid.

#### Part 4:
```{r,echo=F}
#Question 4
lm.log.income<-lm(log(expenditure)~income,data = expend.df)
tidy(lm.log.income)[,1:3]%>%
  kable(escape = F)%>%
  kable_styling(font_size = 15)
par(mfrow=c(2,2))
plot(lm.log.income)
# #QQ plot much better but not sure about the residual vs fitted if the variance decreases as fitted value increases
```
Fitting the same model as in task 2 but with the response of expenditure logged $log(y_{i})=\beta_{0}+\beta_{1}x_{i}+\epsilon$ gives the above coefficients. The data appears to be heteroscedastic as in the fitted vs residual plot the variance in the residuals decreases as the fitted values increases, furthermore the linear assumption also appears violated as the residuals appear to exhibit a slight negative quadratic shape. The QQ plot shows the residuals stay close to the straight line thus normality assumption is valid. 

#### Part 5:
```{r,echo=F}
lm.log.income.sqd <- lm(log(expenditure)~income+I(income^2),data = expend.df)
tidy(lm.log.income.sqd)[,1:3]%>%
  kable(escape = F)%>%
  kable_styling(font_size = 15)

par(mfrow=c(2,2))
plot(lm.log.income.sqd)
```
Fitting the model $log(y_{i})=\beta_{0}+\beta_{1}x_{i}+\beta_{2}x^2_{i}+\epsilon$ gives the estimated coefficients above. Viewing the diagnostic plots the fitted vs residuals plot shows a fairly even scattering of points around 0 and a linear trend, however the residuals are bunched to the very right of the plot and the variance of the residuals appears to decrease as the fitted values increase, this could be to important variables being left out of the model (MAY NEED TO CHANGE THIS). The non-constant variance assumption being violated is further backed up by the none straight line in the scale location plot. The QQ plot shows the points stay close to the straight line meaning the normality assumption is valid.

#### Task 6
```{r}
#Comparing AIC
paste("AIC")
AIC(lm.income)
AIC(lm.income.sqd)
AIC(lm.log.income)
AIC(lm.log.income.sqd)

#Comparing adjusted R squared
paste("adj R squared")
glance(lm.income)$adj.r.squared
glance(lm.income.sqd)$adj.r.squared
glance(lm.log.income)$adj.r.squared
glance(lm.log.income.sqd)$adj.r.squared
```

The 4th model $log(y_{i})=\beta_{0}+\beta_{1}x_{i}+\beta_{2}x^2_{i}+\epsilon$ has the lowest AIC value of *1528.062* and the highest adjusted R squared value of *0.4635055*. Furthermore, looking at the diagnostic plots the normality assumption is valid and the fitted vs residual plot displays even scattering of points around 0. Therefore the 4th model is my preferred model to describe the relationship between expenditure and income.


```{r}
tidy(lm.log.income.sqd)

mean(expend.df$income)
```
The log-linear model means that the covarites (income) have a multiplicative effect on the response (expenditure). The squared term of income included in the model means the change in the mean of expenditure now depends on the value of income. Income has a positive association with expenditure with an increase in income causing a percent increase in expenditure. With the initial value of income affecting the magnitude of the percent increase on expenditure. Smaller initial values of income cause greater percent increase on expenditure (due to the squared income term being negative).

SAYS SUMMARISE SO MAY NOT INCLUDE THE INFOMATION BELOW

In particular, a 1 pound increase in household income increases household expenditure by $exp(0.0032-0.0000029x-0.0000015)-1$x100%. Therefore for a household earning the mean income of £512, a £1 increase in income increases expenditure by approximately **0.17%**. Using a more realistic increase in weekly income of £20 for a household earning the mean income of £512 gives an increase of $exp(0.016-0.000015*20-0.000036)-1$x100% = **1.6%**.


#### Task 7
To start I investigated the multicolinearity.

```{r}
#Checking variance inflation factors (generalized as variables include factors)
car::vif(lm(log(expenditure)~income+I(income^2)+hh.size+hh.adults+sex.hh+lab.force+house.ten,data=expend.df))
```
None of the variables had a GVIF greater than 5, thus multicolinearity is not an issue with the explanatory variables. However, considering the variables it seems likely that household size is related to number of household adults.

```{r}
#Creating tables for chi-squared test
prop.table(table(expend.df$hh.size,expend.df$hh.adults),2)
```
It is clear there is a relationship between the two from the above proportional table. I then created linear models including each separately and both together.

```{r}
lm.adult <- lm(log(expenditure)~income+I(income^2)+hh.adults,data = expend.df)
summary(lm.adult)
#adult significant
  
lm.size <- lm(log(expenditure)~income+I(income^2)+hh.size,data = expend.df)
summary(lm.size)
#size significant

lm.adult.size <- lm(log(expenditure)~income+I(income^2)+hh.size+hh.adults,data = expend.df)
summary(lm.adult.size)
#adult no longer significant
```

```{r}
# #Checking which of the correlated variables to include
# lm.hh.size <- lm(log(expenditure)~income+I(income^2)+hh.size, data = expend.df)
# lm.hh.adults <- lm(log(expenditure)~income+I(income^2)+hh.adults, data = expend.df)
# 
# AIC(lm.hh.size)
# AIC(lm.hh.adults)
# #size has lower AIC
# 
# glance(lm.hh.size)$r.squared
# glance(lm.hh.adults)$r.squared
# #Size has higher R^2

#Based on AIC and R^2 (can use as all models have same number of variables) I select household size and tenure as the 2 variables to use in the model and won't use labour or household adults variables so will no longer consider these 2 variables.
```
Looking at the summary outputs the number of adults in a household is a significant predictor when on it is the only variable added to chosen model. However, when we add the size of a household variable aswell it is no longer significant. Likely the household size was is confounding the effect of number of adults. Thus, I will include the size of a household in for the final model but not the number of adults. 

Considering the possible interactions, I decided to only include 2 way interactions as higher order interactions make interpretation of the model much more complicated and there is no clear 3 way interactions that make sense.

First I investigated interactions between income and the factor variables. To do this I created scatter plots of income against expenditure with income split into each factor of the chosen explanatory factor variable with a fitted linear model line plotted for each of the levels of the factor. Thus it will be possible to see if the effect of income varies within each level of the factor variable on expenditure. This would be seen by a different gradient of the fitted linear model lines.

```{r}
#interaction between 3 factor variables and income investigation
interaction.continous <-function(var){
  ggplot(data=expend.df,aes(x=income,y=expenditure,colour=.data[[var]]))+
    geom_point()+
  geom_smooth(method='lm', formula= y~x,se=F)
}


int1<-interaction.continous("house.ten")
int2<-interaction.continous("sex.hh")
int3<-interaction.continous("hh.size")
int4<-interaction.continous("lab.force")

grid.arrange(int1,int2,int3,int4,ncol=2,nrow=2)
int1
int2
int3
int4
```
Looking at the plots all show that each of the factor variables have an affect on the effect of income on expenditure with different gradients between the levels, however the magnitude of the difference in these gradients varies. For the variable of sex of the household head the difference in gradient is very small indicating that sex has little affect on the effect of income on expenditure. Labour force however has a large difference in gradients between its levels giving evidence of an interaction between labour force and income. I decided to not include a interaction between sex and income as there is little evidence to support this (and makes logical sense).

To check if the remaining 3 interactions with income are significant I added each separately to the chosen model (aswell as the factor variable which is interacting with income).

Looking at the plots. The largest differences in gradients between levels are in income:labour force, income:household adults and income: household size plots, thus these are the most likely to have interactions with income. Furthermore, it makes logical sense that these variables would affect to effect of income on expenditure. Meanwhile sex has little effect on income with the gradients although crossing remaining very similar and house tenure type having some differences. 

```{r}
lm.int.income.size <- lm(log(expenditure)~income+I(income^2)+hh.size*income+sex.hh+lab.force+house.ten,data=expend.df)
summary(lm.int.income.size)
#significant

lm.int.income.labour <- lm(log(expenditure)~income+I(income^2)+hh.size+sex.hh+lab.force*income+house.ten,data=expend.df)
summary(lm.int.income.labour)
#significant

lm.int.income.tenure <- lm(log(expenditure)~income+I(income^2)+hh.size+sex.hh+lab.force+house.ten*income,data=expend.df)
summary(lm.int.income.tenure)
#significant
```
All 3 interactions were significant so I considered all 3 for my final model.

Considering factor variable interactions. There are 6 possible factor interactions. 

```{r}
lm.int.sex.size <- lm(log(expenditure)~income+I(income^2)+hh.size*sex.hh+lab.force+house.ten,data=expend.df)
summary(lm.int.sex.size)
#significant

lm.int.sex.labour <- lm(log(expenditure)~income+I(income^2)+hh.size+sex.hh*lab.force+house.ten,data=expend.df)
summary(lm.int.sex.labour)
#not significant

lm.int.sex.tenure <- lm(log(expenditure)~income+I(income^2)+hh.size+sex.hh*house.ten+lab.force,data=expend.df)
summary(lm.int.sex.tenure)
#not significant

lm.int.size.labour <- lm(log(expenditure)~income+I(income^2)+hh.size*lab.force+sex.hh+house.ten,data=expend.df)
summary(lm.int.size.labour)
#not significant

lm.int.size.tenure <- lm(log(expenditure)~income+I(income^2)+hh.size*house.ten+sex.hh+lab.force,data=expend.df)
summary(lm.int.size.tenure)
#not significant

lm.int.labour.tenure <- lm(log(expenditure)~income+I(income^2)+hh.size+sex.hh+lab.force*house.ten,data=expend.df)
summary(lm.int.labour.tenure)
#significant
```
Adding each to the chosen model separately (plus the variables which were interacting) only sex:household size and labour:tenure were significant, therefore these are the only 2 I will consider for the final model.

#### From full model
To choose my final model I will start with a 'full' model that includes all variables that I want to consider. Then using AIC and adjusted R^2 as my model selection criteria I will remove explanatory variables (those that cause the AIC to become lower or increase the adjusted R^2 when removed). After this as the model is for exploratory analysis and not prediction a simpler model that is easier to interpret/ understand is preferable (Occam's razor), therefore if the model can simplified to only include the important variables and this does not significantly affect the AIC and R^2 this further reduced model will be my final model.

```{r}
#Full model of all variables I want to consider
lm.log.full<-lm(log(expenditure)~income*lab.force+I(income^2)+income*house.ten+income*hh.size+sex.hh*hh.size+lab.force:house.ten,data=expend.df)

AIC(lm.log.full)

#Iteratively performing model selection process
drop1(lm.log.full)
#removing lab.force:house.ten will decrease AIC
lm.full.1.deletion <- update(lm.log.full,.~.-lab.force:house.ten)
AIC(lm.full.1.deletion)

drop1(lm.full.1.deletion)
#removing any variables increases AIC thus stop iterative process

#comparing with original model
glance(lm.log.income.sqd)$adj.r.squared
glance(lm.full.1.deletion)$adj.r.squared

#AIC has been reduced and adjusted R^2 increased

anova(lm.full.1.deletion)
#All variables left statistically significant except income:house.ten
```

Using AIC as the model selection criteria I removed 1 interactions labour force:house tenure type. Comparing this new model to the original the AIC is significantly lower and the adjusted R squared is significantly higher. All variables left in the model are also statistically significant except for income:house tenure.

I then checked how important (magnitude of impact on the response) the variables left in the model are. To find their impact on expenditure I exponentiated the estimated coefficients and note that due to log-linear model the effect is multiplicative.
```{r}
data.frame("coefficent"=tidy(step.method)[,1],exp(tidy(step.method)[,2]))
```
I note that all income interactions are not meaningful/important as their exponentiated value is approximately 1. Thus, their effect on expenditure is negligible. Thus I remove them from the model as a simpler model is preferred.

```{r}
#updating to simplier model
lm.simple.model.1 <- update(lm.full.1.deletion, .~.-income:lab.force-income:hh.size-income:house.ten)

summary(lm.simple.model.1)

drop1(lm.simple.model.1)
#lowest AIC when don't remove any variables

#Therefore this is my final model
lm.final.model <- lm.simple.model.1
```


```{r}
par(mfrow=c(2,2))
plot(lm.final.model)
```
Diagnostic plots show no issues with the model fit. The residuals display a random scattering of points indicating the linearity assumption has been met and the residuals variance does not appear to increase/decrease except for a few values but I don't think this is enough evidence to say that the homoscedasticity assumption is invalid. The points follow a straight line in the QQ plot with the tails deviating slightly but not extreme enough to indicate the normality assumption is invalid. The cooks distance plot shows no influential points are present in the data. Therefore 

## Task 8

The final model is $log(y_{ijkwz})=\beta_{0}+\beta_{1}x_{ijkwz}+\beta_{2}x_{ijkwz}^{2}+\alpha_{j}^{G}+\alpha_k^{S}+\alpha_{w}^{T}+\alpha_{z}^{L}+\alpha^{SG}_{jk}+\epsilon_{ijkwz}$.

```{r}
data.frame("coefficent"=tidy(lm.final.model)[,1],exp(tidy(lm.final.model)[,2]),tidy(lm.final.model)[,5])


summary(lm.final.model)

#Calculating confidence intervals 

lwr <-exp(tidy(lm.final.model)$estimate+c(-1)*qnorm(0.975)*tidy(lm.final.model)$std.error)
upr <- exp(tidy(lm.final.model)$estimate+c(1)*qnorm(0.975)*tidy(lm.final.model)$std.error)

Point.estimates <- round(exp(tidy(lm.final.model)[,2]),3)
p.values <- round(tidy(lm.final.model)[,5],3)

data.frame("coefficent"=tidy(lm.final.model)[,1],
                 Point.estimates,
                 "Confidence Interval 95%"=c(paste("(",round(lwr,3),", ",round(upr,3),")")),
                 p.values)%>%
  kable(digit=3,caption = "Summary of variables affecting expeniture")%>%
  kable_classic(full_width=F,latex_options = c("hold_position"))
```
Due to the model being a log-linear model the explanatory variables have a multiplicative effect on the response. Exponentiating the estimated coefficients to find their effect on expenditure gives the above table. I note that all values have been rounded to 3 decimal places thus the zero p-values are due to p-values being too small be displayed and similarly the effect of income squared is too small to be shown.

Interpreting the above table the intercept value is for a male headed household of 1 person, publicly rented in full time employment who has an income of £0. Clearly this does not make sense

My inference is drawn from the final model summary [**table x**] above. The difference between full time work and being unemployed was highly statistically significant. With those unemployed having an associated decrease in expenditure of (*8.5%*,*34.1%*) 95% CI compared to those in full time work with all other variables held constant. However the difference between full time work and being inactive or in part time work was not significant (CI included 1) therefore with all other variables constant there is no evidence that being in one of these 3 factors compared to another influences the expenditure of the household.

The difference between publicly rented properties expenditure and either privately rented or owned properties expenditure were statistically significant. With the associated increase in expenditure of (*22%*,*44.1%*) and (*11.5%*,*28.1%*) 95% CI's respectively with all other variables held constant.

Due to squared income squared the effect of an increase in income on expenditure depends on the initial value of income. Assuming all other variables are held constant the percentage increase o expenditure is $(\frac{exp(\beta_1(x+y)+\beta_2(x+y)^2)}{exp(\beta_1x+\beta_2x^2)}-1)*100$ where x is the initial value of income and y is the increase in income, $\beta_1=0.002285$ and $\beta_2=-0.000001036$. Thus, for a household earning the mean income of £512, a £20 increase in income increases expenditure by approximately *2.4%* whereas for a household earning only £300 the same £20 increase, increases expenditure by approximately *3.3%*. Due to the negative quadratic term the higher the initial income the less of a positive multiplicative effect an increase in expenditure has on expenditure.

```{r,echo=F}
#calculating percent increase
beta_1 <- 0.002285
beta_2 <- -0.000001036
x <- 300
inc <- 20

(exp(beta_1*(x+inc)+beta_2*(x+inc)^2)/exp(beta_1*x+beta_2*x^2)-1)*100
```

The final 2 variables, sex of household head and household size have an interaction between them. However, only the household size of 3 level has a statistically significant interaction with sex of household head. Therefore as the other levels 

The sex of the household head is highly statistically significant with being female head increasing the expenditure by (*5.6%*,*25.9%*) 95% CI compared to a male head with all other variables held constant. Household size is extremely statistically significant with an increase in expenditure of (*22.4%*,*45.5*),  (*31.3%*,*66.8%*), (*34.9%*,*81.6%*) 95% CI's for households of size 2,4,5+ when compared to a household of size 1 with all other variables held constant. For the effect on expenditure between a household of size 1 and size 3 because of the significant interaction with sex if the household head is male the CI is (*36.2%*,*71.9%*) and (*-15.7%*,*+52.3%*) if female is it is with all other variables held constant. Note that the for the female head CI 0% is included therefore there is no statistically significant difference in expenditure between a household of size 1 with male head and a household of size 3 with a female head with all other variables held constant.

Overall an increase in income, higher household size, being a female head and living in a owned or privately rented increase the % expenditure when compared to a single male headed household living in a publicly rented house (unless a female head with size 3 household). Whereas being unemployed compared to in full time work decreases the % of expenditure.

All the variables mentioned above were statistically significant in their association with expenditure. However, just because they are significant, it does not mean they influence expenditure. There could be confounding or simple chance causing these results. However, some areas of confidence are:

* income, logically if you make more money you have more available to spend 
* household size, larger households are more expensive thus require more spending

sex, tenure, labour force


